refnx -- Neutron and X-ray reflectometry analysis in Python

### 2018-07-25 paper/manuscript.tex
PAPER: method redux

\newcommand{\refnx}{\emph{refnx}}
\newcommand{\Objective}{\textbf{Objective}}
\newcommand{\CurveFitter}{\textbf{CurveFitter}}
\newcommand{\GlobalObjective}{\textbf{GlobalObjective}}
\newcommand{\Parameter}{\textbf{Parameter}}
\newcommand{\Structure}{\textbf{Structure}}
\newcommand{\Slab}{\textbf{Slab}}
\newcommand{\Component}{\textbf{Component}}
\newcommand{\LipidLeaflet}{\textbf{LipidLeaflet}}
\newcommand{\Transform}{\textbf{Transform}}
\newcommand{\DataD}{\textbf{Data1D}}
\newcommand{\ReflectModel}{\textbf{ReflectModel}}
\newcommand{\CurveFitter}{\textbf{CurveFitter}}
\newcommand{\Spline}{\textbf{Spline}}
\newcommand{\conda}{\emph{conda}}
\newcommand{\corner}{\emph{corner}}
\newcommand{\MixedReflectModel}{\textbf{MixedReflectModel}}
\newcommand{\pip}{\emph{pip}}
\newcommand{\emcee}{\emph{emcee}}
\newcommand{\ptemcee}{\emph{ptemcee}}
\newcommand{\NumPy}{\emph{NumPy}}
\newcommand{\SciPy}{\emph{SciPy}}
\newcommand{\Cython}{\emph{Cython}}
\newcommand{\Jupyter}{\emph{Jupyter}}
\newcommand{\ipywidgets}{\emph{ipywidgets}}

@@

The \Objective\ class uses the \ReflectModel\ and a dataset, \DataD, to calculate $\chi^2$, log-likelihood, log-prior, residuals, and the generative model.
The \DataD\ object has \emph{x, y, y\_err, x\_err} attributes to represent $Q$, $R$, $\mathrm{d}R$, $\mathrm{d}Q$. The \DataD\ object can be constructed from a tuple of arrays representing the data, or by reading a three or four column plain-text datafile. A three column dataset represents $Q$ (\si{\per\angstrom}), $R$, $\mathrm{d}R$ (1 standard deviation). A four column dataset represents $Q$, $R$, $\mathrm{d}R$, $\mathrm{d}Q$ (\si{\per\angstrom}). Subclassing \DataD\ would allow other formats to be read. One example of this could be an wavelength dispersive file using $\Omega$, $\lambda$ data instead of $Q$; such as that used in energy scanned X-ray reflectometry, or sometimes produced by wavelength dispersive neutron reflectometers. In such a case \ReflectModel\ could be subclassed to make full use of this energy dispersive information.
An \Objective\ can be given a \Transform\ object to permit fitting as $\log_{10} R$ vs $Q$, $RQ^4$ vs $Q$; the default (no \Transform) is $R$ vs $Q$. Several \Objective\ can be combined to form a \GlobalObjective\ for co-refinement. The object-oriented nature allows \Parameter\, \Component\ to be reused in in different \Objective, thereby facilitating co-refinement.

The \Objective\ statistics are used directly by the \CurveFitter\ class to perform least-square fitting or Bayesian MCMC of the system. \refnx\ uses the \SciPy\ package for least-squares analysis (DifferentialEvolution, Levenberg--Marquardt, LBFGSB - Limited Broyden--Fletcher--Goldfarb--Shanno with bounds) and the \emcee\ and \ptemcee\ packages \cite{emcee, ptemcee} for affine invariant MCMC ensemble sampling of the posterior distribution of the parameters. \ptemcee\ is a fork of \emcee\ that implements parallel tempering, for characterisation of multi-modal probability distributions; different modes can be traversed by chain populations at higher `temperatures', while individual modes are efficiently explored by chains at lower `temperatures'.
The likelihoods used in MCMC and least-squares assume that the measurement errors are normally distributed, Equation~\ref{eqn:2}. However, other types of measurement errors (e.g. Poissonian) could be implemented by a subclass of \Objective\ overriding the log-likelihood method.
Parallelisation of the sampling automatic, and can use MPI on a cluster for greater parallelisation. Visualisation of the samples produced by MCMC sampling is conveniently performed using the \corner\ package \cite{corner} for scatter plot matrices. On the conclusion of a fit/sampling run each \Parameter\ value is updated and given a standard error. For the sampling, these represent the median and half the $[15.87, 84.13]$ percentile range respectively; the latter approximating the standard deviation for a normally distributed statistic.

To make it easy to create quick models, and to ease the learning curve, an ipywidgets \cite{ipywidgets} based fitting graphical user interface (GUI) can be instantiated within a \Jupyter\ notebook, Figure~\ref{fig:gui}. This GUI builds slab based models for analysing a single dataset, but can be used as a stepping point for building more advanced models via the `To code' code generation button. A fully functional web-based reflectometry analysis notebook is currently available \cite{Nelson2018}. For a comprehensive demonstration of multiple contrast co-refinement, see the annotated notebook in the supporting information.

@@

Neutron reflectometry is an ideal technique for the study of biologically relevant lipid membrane mimics and their interactions with proteins, etc. Multiple contrast variation measurements are necessary to reduce modelling ambiguity (due to loss of phase information in the scattering experiment) and improve the ability to determine the structure of various components in the system. The gold standard approach for analysis of these datasets is co-refinement with a common model, and to parameterise the model in terms of chemically relevant parameters, such as the area per molecule. Sometimes a patchy coverage (distinct to low area per molecule) necessitates the use of an (incoherent) sum of reflectivities from different areas. \refnx\ has functionality for all these requirements, such as the \LipidLeaflet\ component for describing the head and tail groups of a lipid leaflet, and \MixedReflectModel to add reflectivities. It is straightforward to develop/modify new components for different structural functionality, as a consequence of the object-oriented approach.

>>>>

\newcommand{\refnx}{\emph{refnx}}
\newcommand{\Objective}{\textbf{Objective}}
\newcommand{\GlobalObjective}{\textbf{GlobalObjective}}
\newcommand{\Parameter}{\textbf{Parameter}}
\newcommand{\Structure}{\textbf{Structure}}
\newcommand{\Slab}{\textbf{Slab}}
\newcommand{\Component}{\textbf{Component}}
\newcommand{\LipidLeaflet}{\textbf{LipidLeaflet}}
\newcommand{\Transform}{\textbf{Transform}}
\newcommand{\DataD}{\textbf{Data1D}}
\newcommand{\ReflectModel}{\textbf{ReflectModel}}
\newcommand{\CurveFitter}{\textbf{CurveFitter}}
\newcommand{\Spline}{\textbf{Spline}}
\newcommand{\conda}{\emph{conda}}
\newcommand{\corner}{\emph{corner}}
\newcommand{\MixedReflectModel}{\textbf{MixedReflectModel}}
\newcommand{\pip}{\emph{pip}}
\newcommand{\emcee}{\emph{emcee}}
\newcommand{\ptemcee}{\emph{ptemcee}}
\newcommand{\NumPy}{\emph{NumPy}}
\newcommand{\SciPy}{\emph{SciPy}}
\newcommand{\Cython}{\emph{Cython}}
\newcommand{\Jupyter}{\emph{Jupyter}}
\newcommand{\ipywidgets}{\emph{ipywidgets}}

@@

The \Objective\ class uses the \ReflectModel\ and a dataset, \DataD, to calculate $\chi^2$, log-likelihood (Equation~\ref{eqn:2}), log-prior, residuals, and the generative model.
The \DataD\ object has \emph{x, y, y\_err, x\_err} attributes to represent $Q$, $R$, $\mathrm{d}R$, $\mathrm{d}Q$. The \DataD\ object can be constructed from a tuple of arrays representing the data, or by reading a three or four column plain-text datafile. A three column dataset represents $Q$ (\si{\per\angstrom}), $R$, $\mathrm{d}R$ (1 standard deviation). A four column dataset represents $Q$, $R$, $\mathrm{d}R$, $\mathrm{d}Q$ (\si{\per\angstrom}). Subclassing \DataD\ would allow other formats to be read. One example of this could be an wavelength dispersive file using $\Omega$, $\lambda$ data instead of $Q$; such as that used in energy scanned X-ray reflectometry, or sometimes produced by wavelength dispersive neutron reflectometers. In such a case \ReflectModel\ could be subclassed to make full use of this energy dispersive information.
An \Objective\ can be given a \Transform\ object to permit fitting as $\log_{10} R$ vs $Q$, $RQ^4$ vs $Q$; the default (no \Transform) is $R$ vs $Q$. Several \Objective\ can be combined to form a \GlobalObjective\ for co-refinement. The object-oriented nature allows \Parameter, \Component\ to be reused in in different \Objective, thereby facilitating co-refinement. For a comprehensive demonstration of multiple contrast co-refinement, see the annotated notebook in the supporting information.

The \Objective\ statistics are used directly by the \CurveFitter\ class to perform least-square fitting with the functionality provided by the \SciPy\ package (DifferentialEvolution, Levenberg--Marquardt, LBFGSB - Limited Broyden--Fletcher--Goldfarb--Shanno with bounds).
\CurveFitter\ can also perform Bayesian Markov Chain Monte Carlo (MCMC) sampling of the system, examining the posterior probability distribution of the parameters, Equation~\ref{eqn:1}. The posterior distribution is proportional to the product of the prior probability and the likelihood (or the sum of the log-probabilities):
%
\begin{gather} 
\label{eqn:1}\ p(\theta | D, I) \propto p(\theta | I)\times p(D | \theta, I)\\
p(D | \theta, I) = -\frac{1}{2} \sum \left[\left(\frac{y_n - model} {\sigma_n}\right)^2 + \log(\sigma_n^2)\right]\label{eqn:2}
\end{gather}
%
The prior distribution, $p(\theta | I)$, is the probability distribution function for a parameter, $\theta$, given preexisting knowledge of the system, $I$, as outlined above.
The likelihood (Equation~\ref{eqn:2}), $p(D | \theta, I)$, is the probability of the observed data, $D$, given the model parameters and other prior information. It is calculated from the measured data, $y_n$ (with uncertainties $\sigma_n$), and the generative model.

The posterior distribution, $p(\theta | I, D)$, is the probability distribution function of the parameters, given the data and prior knowledge about the system.  The posterior probability is examined by encoding the likelihood and prior distributions and using the \emcee\ and \ptemcee\ packages to perform affine invariant ensemble sampling \cite{emcee, ptemcee}. \ptemcee\ is a fork of \emcee\ that implements parallel tempering for characterisation of multi-modal probability distributions; different modes can be traversed by chain populations at higher `temperatures', while individual modes are efficiently explored by chains at lower `temperatures'.

The likelihoods used in MCMC and least-squares assume that the measurement errors are normally distributed, Equation~\ref{eqn:2}. However, other types of measurement errors (e.g. Poissonian) could be implemented by a subclass of \Objective\ overriding the log-likelihood method.

At the end of an MCMC run the parameter set possesses a number of samples (called a `chain'), whose statistics reveal the distribution and covariance of the parameters, the spread of the model around the data, and in a reflectometry context the range of structures that are consistent with the data.
The chain statistics are used to update each \Parameter\ value, and assign a standard error. For the sampling, these represent the median and half the $[15.87, 84.13]$ percentile range respectively; the latter approximating the standard deviation for a normally distributed statistic.

Parallelisation of the sampling is automatic, and can use MPI on a cluster for greater parallelisation. Visualisation of the samples produced by MCMC sampling is conveniently performed using the \corner\ package for scatter plot matrices \cite{corner}.

To make it easy to create quick models, and to ease the learning curve, a browser-based graphical user interface (GUI) \cite{ipywidgets} can be instantiated for fitting within a \Jupyter\ notebook, Figure~\ref{fig:gui}. This GUI uses slab based models for analysing a single dataset, and can be used as a stepping point for building more advanced models via the `To code' code generation button. A fully functional web-based reflectometry analysis notebook is currently available \cite{Nelson2018}.

@@

Neutron reflectometry is an ideal technique for the study of biologically relevant lipid membrane mimics and their interactions with proteins, etc. Multiple contrast variation measurements are necessary to reduce modelling ambiguity (due to loss of phase information in the scattering experiment) and improve the ability to determine the structure of various components in the system. The gold standard approach for analysis of these datasets is co-refinement with a common model, and to parameterise the model in terms of chemically relevant parameters, such as the area per molecule. Sometimes a patchy coverage (distinct to low area per molecule) necessitates the use of an (incoherent) sum of reflectivities from different areas. \refnx\ has functionality for all these requirements, such as the \LipidLeaflet\ component for describing the head and tail groups of a lipid leaflet, and \MixedReflectModel\ to account for patchiness. It is straightforward to develop/modify new components for different structural functionality, as a consequence of the object-oriented approach.
